---
title: "Differential expression results"
author: "Andreas Gschwind"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# save.image("de_results.rda")
# stop()
```

## Goal
Exploration of differential expression results for sample `r snakemake@wildcards$sample`.

```{r requiredPackages, warning=FALSE, message=FALSE}
library(rtracklayer)
library(tidyverse)
library(cowplot)
library(ggrepel)
library(knitr)
```

***

## CRE-level perturbations
CRISPRi perturbations are collapsed per targeted CRE, i.e. any cell expressing a guide targeting a
given CRE is considered perturbed for this CRE. These CRE-level perturbations are used to test for
CRE-gene links and are the primary output of the differential expression analysis.

```{r}
# load files containing differential expression results
results <- read_tsv(snakemake@input[[1]], show_col_types = FALSE)

# add column indicating significant CRE-gene pairs
fdr_threshold <- 0.05
results <- mutate(results, significant = pval_adj < fdr_threshold)
```

### TSS targeting controls
TSS targeting perturbations were included in the experiment as positive controls. These are used to
assess how well the differential expression testing pipeline is well calibrated.

```{r}
# label for TSS controls
tss_ctrl_label <- "TSSCtrl"

# get TSS targeting controls and add a column specifying if the gene corresponds to the intended 
# target (i.e is the gene of the perturbed TSS)
tss_results <- results %>% 
  filter(target_type == tss_ctrl_label) %>% 
  mutate(intended_target = perturbation == gene)

# get maximum absolute log-fold change
max_lfc <- max(abs(tss_results$logFC), na.rm = TRUE)
```

```{r, fig.height=4, fig.width=6}
# create volcano plot for TSS controls
ggplot(tss_results, aes(x = logFC, y = -log10(pval_adj), color = intended_target)) +
  geom_point(data = filter(tss_results, intended_target == FALSE), alpha = 1) + 
  geom_point(data = filter(tss_results, intended_target == TRUE), alpha = 1) + 
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") + 
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("TRUE" = "steelblue", "FALSE" = "gray42")) +
  scale_x_continuous(limits = c(-max_lfc, max_lfc)) +
  labs(title = "All TSS targeting controls", y = "-log10(FDR)", color = "Intended\ntarget") +
  theme_bw()
```

### CRE - gene pairs
The vast majority of guides are targeting putative enhancers. Differential expression results for
these are extracted to assess whether they seem biologically reasonable. Most perturbations
targeting enhancers are expected to decrease the expression of target genes. (Strong) enhancers are
also expected to be enriched within proximity of their target gene.
```{r, fig.height=3.5, fig.width=10}
# remove any TSS controls and all pairs within 1kb to 1Mb of the TSS
enh_results <- results %>% 
  filter(target_type != tss_ctrl_label) %>%
  filter(abs(dist_to_tss) > 1000, abs(dist_to_tss) <= 1e6)

# get maximum absolute log-fold change
max_lfc <- max(abs(enh_results$logFC), na.rm = TRUE)

# make volcano plot with all pairs
p1 <- ggplot(enh_results, aes(x = logFC, y = -log10(pval_adj), color = significant)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") + 
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("TRUE" = "firebrick3", "FALSE" = "gray42")) +
  scale_x_continuous(limits = c(-max_lfc, max_lfc)) +
  labs(title = "All tested enhancer-gene pairs (<=1Mb)", y = "-log10(FDR)") +
  theme_bw()

# plot -log10 p-value vs distance to TSS
p2 <- ggplot(enh_results, aes(x = abs(dist_to_tss) / 1000, y = -log10(pval_adj),
                              color = significant)) +
  geom_point(data = filter(enh_results, significant == FALSE), alpha = 0.5) +
  geom_point(data = filter(enh_results, significant == TRUE), alpha = 0.5) +
  labs(title = "Significance vs. distance to TSS", x = "Distance to TSS (kb)", y = "-log10(FDR)") +
  scale_color_manual(values = c("TRUE" = "firebrick3", "FALSE" = "gray42")) +
  theme_bw()

plot_grid(p1, p2, nrow = 1)
```
