---
title: "FlowFISH guide variability"
author: "Andreas Gschwind"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Goal
Investigate guide to guide variability in FlowFISH data and see if this could be used for a better
estimate of guide variability for power simulations.

```{r requiredPackages, warning=FALSE, message=FALSE}
library(tidyverse)
library(here)
library(cowplot)
library(MASS)
```

***

## Used data
Fulco et al FlowFISH data provided by Ben is used for this analysis.
```{r}
# directory containing data
data_dir <- "/scratch/users/bgrd/share/180928_AllScreens_75TSS"

# get directories containing data without replicates for each gene
gene_dirs <- grep(list.dirs(data_dir, recursive = FALSE), pattern = "\\.|-", invert = TRUE,
                  value = TRUE)

# get all required guide level and element level files within those directories
element_files <- list.files(path = gene_dirs, pattern = ".FullEnhancerScore.txt", full.names = TRUE)
guide_files <- list.files(path = gene_dirs, pattern = ".collapse.bed$", full.names = TRUE)

# set names of files to gene
names(element_files) <- basename(gene_dirs)
names(guide_files) <- basename(gene_dirs)

# load all input files
per_element <- lapply(element_files, FUN = read_tsv, show_col_types = FALSE)
per_guide <- lapply(guide_files, FUN = read_tsv, show_col_types = FALSE,
                    col_names = c("chr", "start", "end", "guide_effect"))

# combine into 2 data frames
per_element <- bind_rows(per_element, .id = "gene")
per_guide <-  bind_rows(per_guide, .id = "gene")
```

```{r, fig.width=5, fig.height=4}
# add per guide effects to per element data
data <- left_join(per_element, per_guide, by = c("gene", "chr", "start", "end"))

# unnest guide effects and transform to long format
data <- data %>% 
  mutate(guide_effect = strsplit(guide_effect, split = ",")) %>% 
  unnest(cols = guide_effect) %>% 
  mutate(guide_effect = as.numeric(guide_effect))

# convert effect sizes to percent changes like in other analyses
data <- data %>% 
  mutate(enh_pct_change = -1 * (1 - mean),
         guide_pct_change = -1 * (1 - guide_effect))

# add unique identified for each tested enhancer - gene pair
data <- unite(data, col = "pair_uid", gene, chr, start, end, remove = FALSE)

# count and plot the number of guides per enhancer - gene pair
guides_per_pair <- count(data, gene, chr, start, end, name = "guides")
ggplot(guides_per_pair, aes(guides)) +
  geom_histogram(binwidth = 1) +
  labs(title = "Number of guides per tested ehancer - gene pair") +
  theme_bw()
```

```{r, fig.width=5, fig.height=4}
# calculate enhancer size per pair
guides_per_pair <- mutate(guides_per_pair, element_width = end - start)

# plot element width vs number of guides
ggplot(guides_per_pair, aes(x = element_width, y = guides)) +
  geom_point(alpha = 0.25) +
  labs(x = "element width (bp)", title = "Number of guides vs. targeted element size") +
  theme_bw()
```

## Enhancer effect size vs guide effect sizes
Only significant enhancers with at least 50 guides are selected and guide-level effects are plotted
against enhancer-level effects.
```{r}
# add the number of guides per pair to data
data <- add_count(data, pair_uid, name = "guides")

# get significant pairs with at least 50 guides targeting the enhancers
sig_pairs_50_guides <- filter(data, Significant == TRUE, guides >= 50, enh_pct_change < 0)

# plot enhancer vs guide effects
p1 <- ggplot(sig_pairs_50_guides, aes(x = enh_pct_change, y = guide_pct_change)) +
  geom_point(alpha = 0.25) +
  geom_smooth(method = "lm", se = FALSE, formula = y ~ x) +
  labs(x = "per-target effect size (pct change)", y = "per-guide effect size (pct change)",
       title = "Per-target vs per-guide effect size") +
  theme_bw() +
  theme(text = element_text(size = 10))
```

```{r}
# fit linear model, extract residuals and add estimated guide variability to data
model <- lm(guide_pct_change ~ enh_pct_change, data = sig_pairs_50_guides)
sig_pairs_50_guides$guide_var <- residuals(model)

# fit normal distribution and extract fitted parameters
distr_fit <- fitdistr(sig_pairs_50_guides$guide_var, densfun = "normal")
distr_param <- distr_fit$estimate 

# plot real guide variability distribution and overlay fitted normal distribution
p2 <- ggplot(sig_pairs_50_guides, aes(guide_var)) +
  geom_histogram(aes(y =..density..), bins = 100) +
  labs(title = "Guide variability distribution",
      x = "Guide effect size normalized for target effect") +
  stat_function(fun = dnorm, args = list(mean = distr_param[["mean"]], sd = distr_param[["sd"]]),
               color = "firebrick3", lwd = 1) +
  scale_x_continuous(breaks = seq(-2, 2, by = 0.25)) +
  theme_bw() +
  theme(text = element_text(size = 10))
```

```{r, fig.width=8, fig.height=4}
plot_grid(p1, p2, rel_widths = c(4, 3))
```

Fitting a normal distribution on the inferred guide variabilities results in a standard distribuion
of `r round(distr_param[["sd"]], digits = 3)`.

## Open questions

* How does guide design affect guide variability? Picking 2-4 best guides vs. all guides.
